\emph{WebAssembly} is a relatively new language, introduced to improve the
performance of compute-intensive workloads in web-based applications. It offers
a compact binary bytecode intended to allow for fast compilation and improved
optimization opportunities over dynamic web languages like JavaScript. These
properties, however, also make it an interesting target for static execution,
enabling web code to run outside of a browser as well as within it. In this
thesis, we describe \emph{SableWasm}, a static, multi-pass compiler system that
translates sandboxed WebAssembly applications to native shared libraries. Our
work covers several different aspects of compiler design. First, we provide an
efficient and extensible WebAssembly module parsing and validation framework,
with improved execution speed and memory footprint compared to the reference
baseline. We then define a middle-level intermediate representation and build
an analysis and transformation framework. We explore several classic data-flow
analyses, such as dominator-tree construction within the framework, and
additionally identify several WebAssembly specific optimization opportunities,
which we address through custom transformation passes. SableWasm also
incorporates several in-progress extension proposals including the SIMD vector
operation extension. Optimized intermediate code is then converted to native
code through a backend implementation with the help of the LLVM compiler
framework and a runtime that enables C/C++ programs to interact with the
WebAssembly module directly. Finally, we evaluate SableWasm by benchmarking
against several well-known testing suites and observe performance improvement
compared to the baseline implementation.