\subsection{Type Inference}
\label{section:mir-opt-type-inference}

This section presents the type system for SableWasm MIR. SableWasm MIR is a
statically typed language with a pretty straightforward type system. However,
one may already notice that SableWasm MIR does not annotate every instruction
with a type, unlike many other compiler intermediate representations. Instead,
SableWasm computes the types for values on-demand via a set of type inference
rules. The type system for SableWasm MIR generalizes from the MVP WebAssembly
type system and its extension proposals with a few modifications. The
formal definition for SableWasm MIR types are as follow,

\begin{lstlisting}[basicstyle=\linespread{1}\ttfamily, mathescape=true]

$\langle$primitive_type$\rangle$ ::= i32 | i64 | f32 | f64 | v128
$\langle$tuple_type$\rangle$     ::= (N, $\langle$primitive_type$\rangle$$\dots$)
$\langle$type$\rangle$           ::= $\langle$primitive_type$\rangle$ | $\langle$tuple_type$\rangle$ | () | $\bot$

\end{lstlisting}

Here we will skip the discussion for \emph{primitive type} and the type checking
rules for its corresponding instructions as they are equivalent to the MVP
WebAssembly type system. The \emph{tuple type} consists of an unsigned integer
and a list of primitive types. They model the return types of multi-value return
functions or \texttt{Pack} instructions. Finally, we introduce the unit type,
$()$, and the bottom type, $\bot$. One can consider the unit type as
\texttt{void} in the C programming language. It represents no value present,
but the type is valid. On the other hand, the bottom type, $\bot$, signals that
the pass cannot assign any valid type to the term. In the rest of this section,
we will focus our discussion on extensions made due to two major WebAssembly
extension proposals, multi-value and SIMD operation.

\paragraph{Multi-value}
WebAssembly multi-value extensions allow functions to have more than one return
values, which is quite interesting. Usually, low-level bytecode representation
does not directly support this feature, and it usually only appears in
higher-level language designs, such as Python. In
section~\ref{section:mir-design-insts}, we introduced
two instructions \texttt{Pack} and \texttt{Unpack}, along with how we represent
multi-value for functions. As a quick recap, SableWasm uses tuples to denote the
multi-value return for functions. The \texttt{Pack} instruction collects values
and constructs a tuple containing them, while on the other hand, the
\texttt{Unpack} extracts primitive values from tuples. Let's focus on the
\texttt{Pack} instruction first. The typing rule for \texttt{Pack} is
straightforward. If we can infer types for all candidate values, we say that the
\texttt{Unpack} instruction has a tuple type consisting of the number of
candidate values and a list of element types. On the other hand, if any of the
candidate values result in a non-primitive type, the \texttt{Pack} instruction
is the $\bot$ type. More formally,
$$
    \frac
    {\Gamma \vdash v_0 \Rightarrow t_0, \dots, v_n \Rightarrow t_n \qquad \forall i, t_i \in primitives}
    {\Gamma \vdash \text{\textbf{pack} } v_0, \dots, v_n \Rightarrow \langle n, t_0 \dots t_n \rangle}
    \qquad
    \frac
    {\Gamma \vdash \exists i, v_t \notin primitives}
    {\Gamma \vdash \text{\textbf{pack} } v_0, \dots, v_n \Rightarrow \bot}
$$
Here the set $primitives$ is the set of all possible primitive types in the
SableWasm MIR type system. For \texttt{Unpack} instructions, the type checker
will first check if the immediate index is within the tuple size. If the index
is out of bounds, the type checker will assign the instruction with bottom type
$\bot$. Otherwise, it will take the type from the tuple specified by the index.
Formally,
$$
    \frac
    {\Gamma \vdash v \Rightarrow \langle n, t_0 \dots t_n \rangle \qquad 0 \leq k \leq n}
    {\Gamma \vdash \text{\textbf{unpack } k v} \Rightarrow t_k}
    \qquad
    \frac
    {\Gamma \vdash v \Rightarrow \langle n, t_0 \dots t_n \rangle \qquad \text{otherwise}}
    {\Gamma \vdash \text{\textbf{unpack } k v} \Rightarrow \bot}
$$
We also generalize the function type in WebAssembly so that SableWasm MIR's
function type will always have a single return value. We use the following
strategy to map WebAssembly's function type into SableWasm MIR function type.
In the case where there are no return values, we translate the return type into
unit type. For example, SableWasm translate \texttt{[i32] -> []} into
\texttt{[i32] -> ()}. On the other hand, if the function type has exactly one
return value, the translation rule is trivial. Finally, when there are multiple
return values, we pack them into a single tuple. For example, SableWasm use
\texttt{[i32] -> (2, i32, f32)} to represent \texttt{[i32] -> [i32, f32]} in
WebAssembly.

\paragraph{SIMD operations}
Section~\ref{section:mir-design-insts} presented the instruction design in
SableWasm MIR. We mentioned
that WebAssembly's 128-bit vector value, added by the SIMD operation extension
proposal, does not store their shape information in the type. WebAssembly's
design gives us two choices in SableWasm when designing a type system for vector
operations. First, we can erase all the shape information for values and
carefully plan the instruction semantics to ensure that all the operations
have defined behaviour at runtime. Second, another approach is to add shape
information back to the values' types. If there is a mismatch in shape
information, either the translation visitor can insert a bit cast, or the type
checker can reject the program. In SableWasm MIR, we take the first approach by
erasing all the shape information from the vector values.
Chapter~\ref{chapter:backend-and-runtime} will introduce the second approach in
detail. The semantics for SIMD instructions in SableWasm MIR follows the
WebAssembly's specification. We always store the value using the little-endian
method and the vectors start their first lane from the least significant bit.

In this section, we talked about the type inference pass in SableWasm MIR.
Similar to the dominator analysis we seen in
section~\ref{section:mir-opt-dominator}, the type infer pass
does not optimize the control flow graph. But they are critical in the backend
when we lower the SableWasm MIR into LLVM. We will come back to this in detail
in chapter~\ref{chapter:backend-and-runtime}.